---
title: "Bike_Sharing"
author: "Charlene"
date: "2023-08-11"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import CSV files
2019 Q1 - 2019 Q4
```{r import file}
library(readr)
library(here)
Q1_2019 <- read.csv(here("data","Divvy_Trips_2019_Q1.csv.gz"))
Q2_2019 <- read.csv(here("data","Divvy_Trips_2019_Q2.csv.gz"))
Q3_2019 <- read.csv(here("data","Divvy_Trips_2019_Q3.csv.gz"))
Q4_2019 <- read.csv(here("data","Divvy_Trips_2019_Q4.csv.gz"))
```
NOTE: Dataset is too large therefore uploading a compressed file 
## Preview data
```{r}
print("column names Q1")
colnames(Q1_2019)
print("column names Q2")
colnames(Q2_2019)
print("column names Q3")
colnames(Q3_2019)
print("column names Q4")
colnames(Q4_2019)
```

## Standardise column name of Q2_2019
Q2 has different column names compared to the rest, so we have to fix that. 
```{r}
library(dplyr)
Q2_2019 <- Q2_2019 %>% 
  rename(trip_id = X01...Rental.Details.Rental.ID,
         start_time = X01...Rental.Details.Local.Start.Time,
         end_time = X01...Rental.Details.Local.End.Time,
         bikeid = X01...Rental.Details.Bike.ID,
         tripduration = X01...Rental.Details.Duration.In.Seconds.Uncapped,
         from_station_id = X03...Rental.Start.Station.ID,
         from_station_name =X03...Rental.Start.Station.Name,
         to_station_id = X02...Rental.End.Station.ID,
         to_station_name = X02...Rental.End.Station.Name,
         usertype = User.Type,
         gender = Member.Gender,
         birthyear = X05...Member.Details.Member.Birthday.Year)
```

## Combining all the 4 quaters to a full year data 
```{r}
data_2019 <- rbind(Q1_2019, Q2_2019, Q3_2019, Q4_2019)
```

## Data Cleaning & Transformation
In order to proceed with our analysis, we need to have two additional information:
1. Ride Duration (end time - start time)
2. Day of week of bike rental

Getting ride duration for each trip:
```{r}
# check if there is any rows with start_time later than end_time:
sum(data_2019$start_time > data_2019$end_time) # 13 rows affected
which(data_2019$start_time > data_2019$end_time) #identify which rows affected
```
To fix affected rows so start time will be later than end time:
```{r}
#change data type to dttm format + create start and end column that satisfy condition start is earlier than end. 
library(dplyr) 
library(lubridate)
data_2019<- data_2019 %>%
  mutate(start = if_else(as_datetime(start_time) > as_datetime(end_time), as_datetime(end_time), as_datetime(start_time)), end = if_else(as_datetime(start_time) > as_datetime(end_time), as_datetime(start_time), as_datetime(end_time))) 

# double check if any start is later than end 
sum(data_2019$start > data_2019$end) 
```

Create a new column to show ride duration:
```{r}
data_2019 <- data_2019 %>% 
  mutate(ride_duration = end - start)

```

Create a new column to show day of week of bike rental
```{r}
data_2019$day = strftime(data_2019$start_time,'%a') #%A means full weekday name
```

## Descriptive Analysis
Average, min and max ride duration (mins):
```{r}
data_2019 %>% 
  group_by(usertype) %>% 
  summarise(mean(ride_duration),max(ride_duration),min(ride_duration))
```

Total number of trips made by Customer VS Subscriber:
```{r}
data_2019 %>% 
  group_by(usertype) %>%
  summarise(Number_customer= sum(usertype == "Customer"),
            Number_subscriber = sum(usertype == "Subscriber")) 
```

Number of trips made by Customer VS Subscriber in each day of the week:
```{r}
data_2019 %>% 
  group_by(usertype,day) %>% 
  summarise(n_distinct(trip_id)) %>% 
  arrange(day)
```

Average ride duration each day of the week Customer VS Subscriber:
```{r}
data_2019 %>% 
  group_by(usertype,day) %>%
  summarise(mean(ride_duration)) %>% 
  arrange(day)
```

## Data Viz

Bike usage trend over the week
```{r}
library(ggplot2)
days_of_the_week <- c("Sun", "Mon", "Tue","Wed","Thu","Fri","Sat")

ggplot(data = data_2019) +
  geom_bar(mapping=aes(x= factor (day,days_of_the_week),fill= usertype), width=.5, position = "dodge")+ labs(title="Bike Usage Trend over the week", x= "Day of Week", y= "Number of Rides", subtitle = "2019") + scale_y_continuous(labels = scales::comma)
```
Avg bike rental duration trend over the week
```{r}
data_2019 %>% 
  group_by(usertype,day) %>%
  summarise(avg_duration = mean(ride_duration)) %>% 
  arrange(day) %>% 
  ggplot(aes(x= factor(day,days_of_the_week),y=avg_duration,fill= usertype)) +
  geom_bar(width=.5, position = "dodge", stat="identity")+ labs(title="Avg Ride Duration Trend over the week", x= "Day of Week", y= "Avgerage Ride Duration (mins)", subtitle = "2019")
```
## Insights 
Based on the Bike usage bar chart, it shows that the usage trend for Customers is opposite that of Subscribers where we are seeing higher usage on weekends for Customers and weekdays for Subscribers.

When looking into the average ride duration of each type of bike users, most Customer rides for at least 50 mins each trip regardless of which day of the week. While, most Subscriber rides between 10 - 20 mins each trip regardless which day of the week.

## Recommendation 
To increase revenue of the bike rental service, management can look into ways to attract new customers by having promotional bundle for longer trips such as: 
1. Full day ride pass on weekends for a fixed price.
2. Half day ride pass for a fixed price.

Additional promotions for weekday bundle to attract more customers to use the rental service on weekdays: 
1. Ride for 40 mins and get an additional 20 mins for free 

Since most subscribers uses the bike rental for short trips (< 20 mins), they are most likely using it as a mode of transport to get from one place to another (eg: work to office). Management can come up with promotional bundles for peak hours such as: lunch hour, morning & evening rush hour when people are commuting to/from work. This will help to give more incentives for users to subscribe and be a member of the bike rental.
